A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-12 tex/sec6-formalisation.tex
[ tex ] Clean up Formalisation

\paragraph{Revisiting language formalisation frameworks}
Unlike prior frameworks~\citep{Allais2021,Fiore2022,Ahrens2022} that have primarily focused on meta-properties centred around substitution for intrinsically-typed terms, our theory of bidirectional type synthesis does not require term substitution but structural induction for extrinsically-typed terms.
The formal definitions of extrinsic typing are more complex than their intrinsic counterparts.
For a specific language such as \PCF, \citet{Wadler2022} noted that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
For our generic definitions, take the formal type-theoretic definition of typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ as an example.
Its intrinsic definition consists of of just one family of sets of intrinsically-typed terms indexed by $A$ and $\Gamma$, but its extrinsic counterpart is a generic family of sets indexed by additionally a generic raw term $t$, involving constructions of two different layers.

\paragraph{Category-theoretic analysis of well-typed terms}
The difference between intrinsic typing and extrinsic typing could be partly manifested by its categorical analysis.
\citet{Fiore1999}{'s} theory of abstract syntax and variable binding forms the foundation of \citet{Fiore2022}'s framework and inspires frameworks by \citet{Allais2021,Ahrens2022}.
Let us take a closer look of their idea.
The set of (untyped) abstract syntax trees for a language can be understood as
\begin{enumerate*}
  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, with
  \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts and the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs. 
  The initiality amounts to structural recursion on terms, namely \emph{term traversal}.
\end{enumerate*}
Substitution is modelled as a monoid multiplication with the notion of strength and a suitable monoidal structure on the category of presheaves.
To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.

Such an initial algebra can be constructed as the least fixpoint of a sequence of objects defined via iteratively applying the signature functor by the Initial Algebra Theorem~\cite{Trnkova1975}.

@@

\begin{figure}
  \codefigure
  \small
  \begin{minipage}[t]{.45\textwidth}
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{decEq}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{DecEq}\;\Conid{Op}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{Op}\;\Varid{→}\;\Conid{ℕ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.45\textwidth}
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;\Conid{SigD}\;\Varid{→}\;\Conid{Set}\;\Varid{→}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Conid{X}\;\hat{}\;(\Conid{D}\;\Varid{.ar}\;\Varid{i}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\Conid{Ty}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Conid{Ξ}\;{}\<[24]%
\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{op}\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;(\Conid{Ty}\;\Conid{Ξ})\;{}\<[24]%
\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Simple types in \Agda}
  \label{fig:formal-simple-type}
\end{figure}
We start with the formal definition of signatures and simple types (\Cref{fig:formal-simple-type}), which parallels its informal counterpart (\Cref{def:simple-signature}), with the exception of \ensuremath{\Varid{⟦\char95 ⟧}}.
To prevent circular references during the inductive definition of simple types, we initially define \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} on an arbitrary \ensuremath{\Conid{Set}} instead of \ensuremath{\Conid{Ty}}.
Thus, types specified by a signature can be defined using two constructors--- \ensuremath{\text{\textasciigrave}\anonymous }\ and \ensuremath{\Varid{op}}---where a variable is represented by \ensuremath{\text{\textasciigrave}\;\Varid{n}} for some \ensuremath{\Varid{n}} less than \ensuremath{\Conid{Ξ}}, and each inhabitant \ensuremath{\Varid{op}\;(\Varid{i}\;\Varid{,}\;\Varid{ts})} is a pair of an operation symbol \ensuremath{\Varid{i}} and a list \ensuremath{\Varid{ts}} of length $n$.

From a categorical standpoint, \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} acts as a functor from \ensuremath{\Conid{Set}} to \ensuremath{\Conid{Set}}, mapping $X$ to an $\Conid{Op}$-indexed coproduct $\sum_{i \in \Conid{Op}} X ^ {\Conid{ar}(i)}$ of products of $X$.
The type \ensuremath{\Conid{Ty}} is the free \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}-algebra over the type \ensuremath{\Conid{Fin}\;\Conid{Ξ}} or the initial algebra for the functor \ensuremath{\Conid{Fin}\;\Conid{Ξ}\;\Varid{+}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}.

Finally, to account for indices when defining raw terms and bidirectional typing derivations by a signature, we define a corresponding signature functor \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} for a (bidirectional) binding signature.

@@

As shown in \Cref{fig:formal-bibisig}, we establish \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\>[3]{}\Keyword{where}{}\<[E]%
\>[3]{}\Keyword{field}{}\<[E]%
\>[5]{}\Varid{cxt}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\>[5]{}\Varid{type}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\>[3]{}\Varid{constructor}\;\Varid{ι}{}\<[E]%
\>[3]{}\Keyword{field}{}\<[E]%
\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\>[5]{}\Varid{type}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
\>[29]{}\Varid{tvars}{}\<[E]%
\>[5]{}\Varid{args}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\>[3]{}\Varid{constructor}\;\Varid{sigd}{}\<[E]%
\>[3]{}\Keyword{field}{}\<[E]%
\>[5]{}\Conid{Op}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\>[5]{}\Varid{ar}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
  \caption{Bidirectional binding signature in \Agda}
  \label{fig:formal-bibisig}

@@

This signature endofunctor for raw term construction resembles the one for defining simple types.
However, we define four distinct functors instead of defining one directly.

@@

Notably, the functor \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} extends the context by $\erase{\Delta}$ variables by mapping the family of sets \ensuremath{\Conid{X}\;\Varid{n}} to \ensuremath{\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n})}.

@@

\subsection{Defining Extrinsic Bidirectional Typing Derivations}\label{subsec:formal-bidirectional-typing}
Formally defining extrinsic derivations is arguably the most complex aspect of our formalisation.
This requires defining a family of sets, indexed by context, type, mode, and notably, a raw term.
This definition involves another signature endofunctor, requiring consistency with the functor used for constructing raw terms.

@@

In a similar vein, we define four functors for the construction of bidirectional typing derivations, albeit in a top-down manner.
For an endofunctor on predicates, each functor must act as a \emph{lifting} of the functor on its domain, illustrated below.
The signature endofunctor \ensuremath{\Varid{⟦\char95 ⟧}} for a bidirectional binding signature is defined as follows:
{\small
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[6]%
\>[6]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{SigD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{o}\;\Varid{,}\;\Varid{ts})\;\Varid{d}\;\Conid{A}\;\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{ts}\;\Varid{d}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}
This functor maps a predicate \ensuremath{\Conid{Y}} over \ensuremath{\Conid{X}} into a predicate \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}} over \ensuremath{\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}}.
Here, \ensuremath{\Varid{erase}\;\Conid{D}} represents the mode erasure of \ensuremath{\Conid{D}}, and the prefix \ensuremath{\Conid{R}} in \ensuremath{\Conid{R.⟦\char95 ⟧}} indicates the signature endofunctor used in raw term construction.

>>>>

\paragraph{Revisiting language formalisation frameworks}
Unlike prior frameworks~\citep{Allais2021,Fiore2022,Ahrens2022} that have primarily focused on meta-properties centred around substitution for intrinsically-typed terms, our theory of bidirectional type synthesis does not require term substitution but structural induction for extrinsically-typed terms.
The formal definitions of extrinsic typing are more complex than their intrinsic counterparts.
%For a specific language such as \PCF, \citet{Wadler2022} noted that `extrinsically-typed terms require about 1.6 times as much code as intrinsically-typed' for their formalisation of type safety.
%For our generic definitions,
Take our formal definition of typing derivations $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ as an example.
Its intrinsic definition is just one family of sets of typed terms indexed by $A$ and $\Gamma$, but its extrinsic counterpart is a generic family of sets indexed by additionally a generic raw term $t$, involving constructions of two different layers.

\paragraph{Category-theoretic analysis of well-typed terms}
The difference between intrinsic typing and extrinsic typing could be partly manifested by \citet{Fiore1999}{'s} theory of abstract syntax and variable binding which forms the foundation of \citet{Fiore2022}'s framework and inspires other referenced frameworks.
Let us take a closer look of their idea.
The set of (untyped) abstract syntax trees for a language can be understood as
\begin{enumerate*}
  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, with
  \item an \emph{initial} algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$ of variables to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts and the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the binding arities of constructs. 
  The initiality amounts to structural recursion on terms, namely \emph{term traversal}.
\end{enumerate*}
Substitution for $\Term$ is modelled as a monoid in the category of presheaves.
To put it succinctly, it is the free $\mathbb{\Sigma}$-monoid over the presheaf~$V$ of variables.

Such an initial algebra can be constructed by the Initial Algebra Theorem~\cite{Trnkova1975}.

@@

\begin{figure}
  \codefigure
  \small
  \begin{minipage}[t]{.45\textwidth}
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{decEq}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{DecEq}\;\Conid{Op}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[13]%
\>[13]{}\mathbin{:}\;{}\<[16]%
\>[16]{}\Conid{Op}\;\Varid{→}\;\Conid{ℕ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.45\textwidth}
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{8}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧}\;\mathbin{:}\;\Conid{SigD}\;\Varid{→}\;\Conid{Set}\;\Varid{→}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\mathrel{=}\;\Conid{Σ[}\;\Varid{i}\;\Varid{∈}\;\Conid{D}\;\Varid{.Op}\;\mskip1.5mu]\;\Conid{X}\;\hat{}\;(\Conid{D}\;\Varid{.ar}\;\Varid{i}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Keyword{data}\;\Conid{Ty}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\text{\textasciigrave}\anonymous \;{}\<[8]%
\>[8]{}\mathbin{:}\;\Conid{Fin}\;\Conid{Ξ}\;{}\<[24]%
\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{op}\;{}\<[8]%
\>[8]{}\mathbin{:}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;(\Conid{Ty}\;\Conid{Ξ})\;{}\<[24]%
\>[24]{}\Varid{→}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Simple types in \Agda}
  \label{fig:formal-simple-type}
\end{figure}
We start with the formal definition of signatures and simple types (\Cref{fig:formal-simple-type}), which parallels its informal counterpart (\Cref{def:simple-signature}), with the exception of \ensuremath{\Varid{⟦\char95 ⟧}}.
To prevent circular references during the inductive definition of simple types, we initially define \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} on an arbitrary \ensuremath{\Conid{Set}} instead of\/ \ensuremath{\Conid{Ty}}.
Thus, types specified by a signature can be defined inductively where each inhabitant \ensuremath{\Varid{op}\;(\Varid{i}\;\Varid{,}\;\Varid{ts})} for the $\Rule{Op}$ rule is a pair of an operation symbol and a list of length $n$.

From a categorical view, \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}} is a functor from \ensuremath{\Conid{Set}} to \ensuremath{\Conid{Set}}, mapping $X$ to an $\Conid{Op}$-indexed coproduct $\sum_{i \in \Conid{Op}} X ^ {\Conid{ar}(i)}$ of products of $X$.
The type \ensuremath{\Conid{Ty}} is the free \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}-algebra over the type \ensuremath{\Conid{Fin}\;\Conid{Ξ}} or the initial algebra for the functor \ensuremath{\Conid{Fin}\;\Conid{Ξ}\;\Varid{+}\;\Varid{⟦}\;\Conid{D}\;\Varid{⟧}}.

@@

As shown in \Cref{fig:formal-bibisig}, we establish \ensuremath{\Conid{ArgD}}, \ensuremath{\Conid{OpD}}, and \ensuremath{\Conid{SigD}} to represent arguments $[\Delta]A^{\dir{d}}$, bidirectional binding operations $\biop$, and bidirectional binding signatures, respectively, in line with their informal definitions.
By removing \ensuremath{\Varid{mode}} from these definitions, we retrieve the definitions of binding arguments, arities, and signatures.
\begin{figure}
  \codefigure
  \begin{minipage}[t]{.3\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{ArgD}\;(\Conid{Ξ}\;\mathbin{:}\;\Conid{ℕ})\;\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{cxt}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Cxt}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Ty}\;\Conid{Ξ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[19]%
\>[19]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Conid{ArgsD}\;\mathrel{=}\;\Conid{List}\;\Varid{∘}\;\Conid{ArgD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.33\textwidth}
  \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{OpD}\;\mathbin{:}\;\Conid{Set}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{ι}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\{\mskip1.5mu \Varid{tvars}\mskip1.5mu\}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ℕ}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\dir{\Varid{mode}}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{Mode}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{type}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{TExp}\;{}\<[29]%
\>[29]{}\Varid{tvars}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{args}\;{}\<[21]%
\>[21]{}\mathbin{:}\;\Conid{ArgsD}\;\Varid{tvars}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \begin{minipage}[t]{.3\textwidth}
   \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Keyword{record}\;\Conid{SigD}\;\mathbin{:}\;\Conid{Set₁}\;\Keyword{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{constructor}\;\Varid{sigd}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Keyword{field}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Op}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Set}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{ar}\;{}\<[15]%
\>[15]{}\mathbin{:}\;\Conid{Op}\;\Varid{→}\;\Conid{OpD}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
  \end{minipage}
  \caption{Bidirectional binding signature in \Agda}
  \label{fig:formal-bibisig}
\end{figure}

@@

This functor for raw term construction resembles the one for defining simple types.
However, we define four distinct functors instead of directly defining one .

@@

Notably, the functor \ensuremath{\Varid{⟦}\;\Conid{Δ}\;\Varid{⟧ᵃ}} extends the context by mapping the family of sets \ensuremath{\Conid{X}\;\Varid{n}} to \ensuremath{\Conid{X}\;(\Varid{length}\;\Conid{Δ}\;\Varid{+}\;\Varid{n})}.

@@

\subsection{Defining Extrinsic Bidirectional Typing Derivations}\label{subsec:formal-bidirectional-typing}
Formally defining extrinsic derivations is much trickier.
This requires defining a family of sets, indexed by context, type, mode, and notably, a raw term, involving another signature endofunctor alone with the functor used for constructing raw terms.

@@

In a similar vein, we define four functors for the construction of bidirectional typing derivations, albeit in a top-down manner.
An endofunctor on predicates must act as a \emph{lifting} of the functor on its domain, illustrated by the signature endofunctor \ensuremath{\Varid{⟦\char95 ⟧}} for a bidirectional binding signature:
{\small
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{⟦\char95 ⟧}\;{}\<[6]%
\>[6]{}\mathbin{:}\;(\Conid{D}\;\mathbin{:}\;\Conid{SigD})\;(\Conid{X}\;\mathbin{:}\;\Conid{R.Fam})\;(\Conid{Y}\;\mathbin{:}\;\Conid{Fam}\;\Conid{X})\;\Varid{→}\;\Conid{Fam}\;(\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}){}\<[E]%
\\
\>[B]{}\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;(\Varid{o}\;\Varid{,}\;\Varid{ts})\;\Varid{d}\;\Conid{A}\;\mathrel{=}\;\Varid{⟦}\;\Conid{D}\;\Varid{.ar}\;\Varid{o}\;\Varid{⟧ᶜ}\;\Conid{X}\;\Conid{Y}\;\Conid{Γ}\;\Varid{ts}\;\Varid{d}\;\Conid{A}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
}
This functor maps a predicate \ensuremath{\Conid{Y}} over \ensuremath{\Conid{X}} into a predicate \ensuremath{\Varid{⟦}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}\;\Conid{Y}} over \ensuremath{\Conid{R.⟦}\;\Varid{erase}\;\Conid{D}\;\Varid{⟧}\;\Conid{X}}.
Here, \ensuremath{\Varid{erase}\;\Conid{D}} represents the mode erasure of \ensuremath{\Conid{D}}, and the prefix \ensuremath{\Conid{R}} in \ensuremath{\Conid{R.⟦\char95 ⟧}} indicates the signature endofunctor used in raw term construction.
